Compositional Quantitative Resource Analysis of C Programs

### 2014-06-13 paper/main.tex
more work on intro

\usepackage{prettyref}
\newcommand{\pref}[1]{\prettyref{#1}}
\newcommand{\Pref}[1]{\prettyref{#1} \vpageref[]{#1}}
\newcommand{\ppref}[1]{\vpageref[]{#1}}
\newrefformat{fig}{Figure~\ref{#1}}
\newrefformat{tab}{Table~\ref{#1}}
\newrefformat{cha}{Challenge~\ref{#1}}
\newrefformat{compiler}{Point~\ref{#1} of \pref{thm:compiler}}

@@

  an experimental evaluation that the our novel amortized resource

@@

Our approach complements existing works since it provides a semantic

@@

The main innovation that makes this work practical is our new
automatic amortized analysis for C code.  It targets programs that are
common in system software, that is, programs whose resource
consumption can be described as a function of sizes $|[x,y]|$ of
intervals of integer variables.  A typical bound would be $B(x,y,z) =
10{\cdot}|[x,y]| + 1{\cdot}|[z,0]|$ and we have for example
$B(100,200,43) = 1043$.
%
Despite the apparent simplicity of the new analysis system, it is able
to reproduce multiple results from the literature which were obtained
using methods based on abstract interpretation.  In contrast with
abstract interpretation--based methods, our technique does not require
any fixpoint computations on an abstract domain to obtain loop
invariants.  The mechanism we designed is able to leverage local
assertions such as $x < y$ that we collect along the branching points
to obtain global invariants.  We achieve this by generating a linear
constraint system that reflects resource cost and size changes of
integer variables in the program.  A solution of the linear program is
a resource usage bound for the C program.  We can naturally, handle
advanced control flow such as break, return, iteration over negative
counter variables, and mutually-recursive functions.

Following the development steps of automatic amortized analysis for
functional programs [XXX], we deliberately restrict our self to linear
bounds in the automatic analysis (there a no restrictions for the
manually-derived bounds in the logic).  More specifically, bounds have
the form $\sum_{a,b} q_{(a,b)} |[a,b]|$ where $a$ and $b$ are integer
variables or constants.  The reason is mainly clarity of presentation,
since an extension to multivariate resource polynomials [XXX] would
make the inference rules considerably more complected and should
better be described separately.  However, we developed the linear
inference system so that the extension to polynomial bounds shall
work smoothly.

@@

Our new quantitative Hoare logic for CompCert Clight is a
generalization of a program logic that we recently developed to
formally very stack-space bounds in Coq [PLDI].  In a nutshell, we
replace the boolean predicates (pre and post conditions) of Hoare
logic with numeric predicates that map a program state to a
non-negative number or infinity.  The intuitive meaning of a
quantitative Hoare triple $\htriple{P}{C}{Q}$ that $C$ consumes $P(S)$
resources when executed in starting state $S$, and that $Q(S')$
resources are left if $C$ terminates in state $S'$.  Infinity
corresponds to \emph{false} in classic Hoare logic.  Our quantitative
logic is proved sound with respect to a cost semantic for Clight that
accounts a constant cost for each evaluation step.  This constant cost
is parametric in a resource metric that can be defined by the user.

We have evaluated the quantitative logic and the automatic analysis
with system code and examples from the literature. XXX

@@

\section{Catalog of Automatically Analyzed Programs}

>>>>

\usepackage{prettyref}
\newcommand{\pref}[1]{\prettyref{#1}}
\newcommand{\Pref}[1]{\prettyref{#1} \vpageref[]{#1}}
\newcommand{\ppref}[1]{\vpageref[]{#1}}
\newrefformat{fig}{Figure~\ref{#1}}
\newrefformat{app}{Appendix~\ref{#1}}
\newrefformat{tab}{Table~\ref{#1}}
\newrefformat{cha}{Challenge~\ref{#1}}
\newrefformat{compiler}{Point~\ref{#1} of \pref{thm:compiler}}

@@

  an experimental evaluation that our novel amortized resource

@@

Our approach complements existing work since it provides a semantic

@@

The main innovation that makes this work practical is our new
automatic amortized analysis for C code.  It targets programs that are
common in system software, that is, programs whose resource
consumption can be described as a function of sizes $|[x,y]|$ of
intervals of integer variables.  A typical bound would be $B(x,y,z) =
10{\cdot}|[x,y]| + 1{\cdot}|[z,0]|$ (for example $B(100,200,43) =
1043$).
%
Despite the apparent simplicity of the new analysis system, it is able
to reproduce results from the literature which were obtained using
methods based on abstract interpretation.  In contrast with abstract
interpretation--based methods, our technique does not require any
fixpoint computations on an abstract domain to obtain loop invariants.
The mechanism we designed is able to leverage local assertions such as
$x < y$ that we collect along the branching points of the program to
obtain global invariants.  We achieve this by generating a linear
constraint system that reflects resource cost and size changes of
integer variables in the program.  A solution of the linear program is
a resource usage bound for the C program.  We can naturally, handle
advanced control flow such as break, return, iteration over negative
integer variables, and mutually-recursive functions.

Following the development steps of automatic amortized analysis for
functional programs [XXX], we deliberately restrict our self to linear
bounds in the automatic analysis (there a no restrictions for the
manually-derived bounds in the logic).  More specifically, bounds have
the form $\sum_{a,b} q_{(a,b)} |[a,b]|$ where $a$ and $b$ are integer
variables or constants.  The reason is mainly clarity of presentation,
since an extension to multivariate resource polynomials [XXX] would
make the inference rules considerably more involved and should
better be described separately.  However, we developed the linear
inference system so that the extension to polynomial bounds shall
work smoothly.

@@

Our new quantitative Hoare logic for CompCert Clight is a
generalization of a program logic that we recently developed to
formally very stack-space bounds in Coq [PLDI].  In a nutshell, we
replace the boolean predicates (pre and post conditions) of Hoare
logic with numeric predicates that map a program state to a
non-negative number or infinity.  The intuitive meaning of a
quantitative Hoare triple $\htriple{P}{C}{Q}$ that $C$ consumes $P(S)$
resources when executed in starting state $S$, and that $Q(S')$
resources are left if $C$ terminates in state $S'$.  Infinity
corresponds to \emph{false} in classic Hoare logic.  Our quantitative
logic is proved sound with respect to a cost semantic for Clight that
accounts a constant cost for each evaluation step.  This constant cost
is parametric in a resource metric that can be defined by the user.
Our shallow embedding in Coq enables the derivation of both, bounds
that are parametric in the cost metric or bounds for one specific
metric.

We have evaluated the quantitative logic and the automatic analysis
with system code and examples from the literature. \pref{app:cat}
contains a $50$ (XXX) challenging loop and recursion patterns that we
collected from open source software and the literature.  Our analysis
can find asymptotically tight bounds for $48$ of these patterns and in
most cases the derived constant factors are tight.  We compared our
automatic analyzer to other resource-analysis systems.  However, our
analyzer and Rank are currently the only publically available tools
that can directly operate on C programs.  We selected $5$ example for
which we derived linear bounds and translated them to term-rewriting
systems by hand in order to run them with the tools KoAT and Pubs.
Both tools and Rank were not able to find linear bounds for the
examples.  We contacted the authors of LOOPUS and SPEED but were not
able to obtain a tool to perform a comparison.

To show the practicability of our approach we developed a case study
with different real-time schedulers implemented in C.  To perform a
schedulability analysis that takes into account the overhead of the
scheduler we derived time bounds for context switching and
initialization of the scheduler.  We then computed time bounds for a
set of given user task and took account the previously computed
scheduler overhead to determine whether the scheduling requirements of
all user tasks can be met.

@@

\section{Catalog of Automatically Analyzed Programs}
\lable{app:cat}
